---
title: "Analysis Code"
author: "Bart Larsen"
date: "10/10/2019"
output:
  html_document: default
  toc: yes
---

This code implements all statistical analyses and creates figures for "Longitudinal development of brain iron is linked to cognition in youth" (Larsen, et al., 2019).

---

```{r setup, include=FALSE}
## Load Libraries
require(ggplot2)
require(gridExtra)
library(cowplot)
require(tidyverse)
require(mgcv)
require(gratia)
require(pbkrtest)
library(broom)
library(kableExtra)
library(parallel)
library(scales)
source('combat.R')
source('utils.R')
font_size <- 16
theme_set(theme_classic(base_family = "sans",base_size = font_size))
line_size <- 1.5
point_size <- 2

```

## Relevant functions:
### GAMM modeling
```{r gamm}
# This function:
# 1. executes the GAMM model,
# 2. sends output to the parametric bootstrap (if requested),
# 3. prints a regression table, and 
# 4. sends the model to the visualizer for plotting.

gamm_model <- function(df, model_formula, this_label, smooth_var, int_var = NULL,group_var, pbootstrap = F, longPlot = F, model_test = T){
  cat(sprintf("\n\n### Results for %s\n",this_label))
  
 model_formula <- as.formula(model_formula)
  if(!"exclusions" %in% colnames(df))  {
    df$exclusions <- FALSE; #there is no exclusions column so make one that excludes none
  }
  g1<-gamm(model_formula,
           data=df,
           random = list(bblid =~ 1),
           subset = exclusions == F)
  
  if (model_test == T){
    if (pbootstrap == T) {
      #Send to bootstrap function
      g1$pb<-pboot(g1) 
      #Print a table that shows the bootstrap outcome
      print(g1$pb %>%
        summary() %>%
        .$test %>%
        as.data.frame() %>%
        kable(caption = sprintf("Parametric Bootstrap test for %s",this_label)) %>%
        kable_styling(full_width = F, position = "left",bootstrap_options = c("striped"))
      )
      
      if (isTRUE(all.equal(g1$pb$bestmod,model_formula))) {
        cat("The initial (more complicated) model is best")
        g <- g1
        
        # Refit the model without fixed df for the purpose of plotting
        plot_formula <- as.formula(gsub(", fx = T","",deparse(model_formula)))
        cat(deparse(plot_formula))
        plotg<-g
        plotg <- gamm(plot_formula,
                      data = df,
                      random = list(bblid =~1),
                      subset = exclusions == F)
      } else {
        cat("The simpler model is best")
        cat("   refitting  ")
        g <-gamm(as.formula(g1$pb$bestmod),
                 data=df,
                 random = list(bblid =~ 1),
                 subset = exclusions == F)
        
        #Again, this part not implemented (model not refit with unfixed df)
        plot_formula <- as.formula(gsub("ti\\(",'te\\(',deparse(g$gam$formula)) %>% gsub(", fx = T", "", .))
        plotg<-g
        plotg <-gamm(plot_formula,
                 data=df,
                 random = list(bblid =~ 1),
                 subset = exclusions == F)
      }
    } else {
      if (!is.null(int_var)) {
        # We are not bootstrapping, but there is an interaction variable
        s<-summary(g1$gam)
        if (s$s.table[grep(x=rownames(s$s.table),pattern = int_var),"p-value"] <.05/4)  {
          #Checked if interaction is sig, if so keep in the model
          g <- g1
          plot_formula <- as.formula(gsub(", fx = T", "", deparse(model_formula)))
          plotg <- g
        } else {
          #Interaction is not sig, remove from the model
          cat("The simpler model is best")
          thisResp <- as.character(g1$gam$terms[[2]])
          theseVars <- attr(terms(model_formula),"term.labels")
          new_formula <- reformulate(theseVars[0:(length(theseVars)-1)],response = thisResp)
          
          g <-gamm(as.formula(new_formula),
                   data=df,
                   random = list(bblid =~ 1),
                   subset = exclusions == F)
          plot_formula <- as.formula(gsub("ti\\(",'te\\(',deparse(g$gam$formula)) %>% gsub(", fx = T", "", .))
          plotg<-gamm(as.formula(plot_formula),
                   data=df,
                   random = list(bblid =~ 1),
                   subset = exclusions == F)
      }
      } else {
        #There is no interaction term, just plot.
        g <- g1
          plot_formula <- as.formula(gsub(", fx = T", "", deparse(model_formula)))
          plotg <-gamm(as.formula(plot_formula),
                   data=df,
                   random = list(bblid =~ 1),
                   subset = exclusions == F)
      }
    }
  } else {
    g <- g1
    plotg<-g
  }

  g$gam$data<-df %>%
    filter(exclusions == F)
  
  #Display model results:
  s_tidytable<- tidy(g$gam)
  p_tidytable <- tidy(g$gam,parametric = T)
  snames = names(s_tidytable)
  pnames = names(p_tidytable)
  names(s_tidytable)=pnames
  thisBIC <- BIC(g$lme)
  numObs <- g$lme$dims$N
  g$BIC <- thisBIC
  print(concurvity(g$gam)%>%kable(caption = "convurvity")%>%kable_styling(full_width = F,bootstrap_options = "striped",position = "left"))
  stattable <- rbind(p_tidytable,snames,s_tidytable) %>%
    kable(caption = sprintf("Regression table from gamm in %s, BIC = %1.2f, obs = %d",this_label,thisBIC,numObs)) %>% 
    kable_styling(full_width = F, position = "left")
  
  print(stattable)
  write.csv(x = rbind(p_tidytable,snames,s_tidytable),file = sprintf("GAMM_table_%s.csv",this_label),row.names = F)
  cat(sprintf("Regression table from gamm in %s, BIC = %1.2f, obs = %d\n",this_label,thisBIC,numObs),
      file = sprintf("GAMM_table_%s.txt",this_label))
  cat(sprintf("Bootstrap p value %1.5f",g1$pb$test["PBtest","p.value"]),
      file = sprintf("GAMM_table_%s.txt",this_label),
      append = T)


  #Send final model to visualizer:
  if (longPlot == T) {
      g$pl <- longitudinal_plot(g,plabels = this_label)
    } else{
    if (s_tidytable$p.value[nrow(s_tidytable)]<1.05) {
        g$pl <- visualize_models(plotg,plabels = this_label, smooth_var = smooth_var, int_var = int_var, group_var = group_var)
    }
    }
  #Return result object
  result <- g
  
  return(result)
}
```

### Bootstrap procedure
```{r include=T,opts.label = "Bootstrap Function"}
## Parametric bootstrap of likelihood ratio test for nested models
pboot <- function(modelobj){
  numsims <- 1000

  df <- modelobj$gam$model
  thisResp <- as.character(modelobj$gam$terms[[2]])
  f1 <- modelobj$gam$formula
  theseVars <- attr(terms(f1),"term.labels")
  f2 <- reformulate(theseVars[0:(length(theseVars)-1)],response = thisResp)
  
  g1 <- gam(f1,data = df)
  g2 <- gam(f2,data = df)

  mat1 <- model.matrix(g1)
  mat2 <- model.matrix(g2)

  bblid<-df$bblid
  y <- df[,thisResp]
  
  m1 <- lmer(y ~ -1 + mat1 + (1|bblid))
  m2 <- lmer(y ~ -1 + mat2 + (1|bblid))
  refdist <- PBrefdist(m1, m2, nsim=numsims)#, cl=cl)
  pb <- PBmodcomp(m1, m2, ref = refdist)
  int_pval <- pb$test["PBtest","p.value"]
  if (int_pval < .05/4) {
    pb$bestmod <- f1
  } else {
    pb$bestmod <- f2
  }
  return(pb)
}
```

### Visualize models
```{r visual func, include=FALSE}
# Func to visualize model outputs
visualize_models <- function(modobj,smooth_var, int_var ,group_var, plabels = NULL,check_diagnostics = F){
  if (any(class(modobj)=="gam")) {
    model <- modobj
  } else if (class(modobj$gam)=="gam") {
    model <- modobj$gam
  } else {
    stop("Can't find a gam object to plot")
  }

  ## Generate custom line plot
  np <- 10000 #number of predicted values
  df = model$model

  theseVars <- attr(model$terms,"term.labels")
  varClasses <- attr(model$terms,"dataClasses")
  thisResp <- as.character(model$terms[[2]])

  if (!is.null(int_var)) {
    if (!any(grepl(x=as.character(model$formula),pattern = int_var))) {
      return()
    }
    switch (varClasses[int_var],
      "numeric" = {
        q <- quantile(df[,int_var],probs = c(.05,.95)) #pick 5% and 95% to plot
        bigq <- q[[2]]
        smallq <- q[[1]]
        values <- c(bigq,smallq)
        labs <- c(sprintf("high (%1.2f)",bigq),sprintf("low (%1.2f)",smallq))

        q <-quantile(rescale(df[,int_var],c(0,1)),probs = c(.05,.5,.95))
        limit_values <- c(q[[1]],q[[length(q)]])
        midpoint_val <- unname(q[[2]])
        
        grad_fill = T
      },
      "factor" = {
        labs <- levels(df[,int_var])
        values <- levels(df[,int_var])
        grad_fill = F
      },
      "ordered" = {
        labs <- levels(df[,int_var])
        values <- levels(df[,int_var])
        grad_fill = F
      }
    )

    labPred <- data.frame(init = rep(0,np*length(labs)))
    labPred[,int_var] = rep(values,each = np)
    labPred$lab = rep(labs,each = np)
    labPred <- labPred[,names(labPred) !="init"]
    thisPred <- data.frame(init = rep(0,np))
    
    for (v in c(1:length(theseVars))) {
      thisVar <- theseVars[[v]]
      thisClass <- varClasses[thisVar]
      if (thisVar == smooth_var) {
        thisPred[,smooth_var] = seq(min(df[,smooth_var],na.rm = T),max(df[,smooth_var],na.rm = T), length.out = np)
      } else if (v == int_var) {
        next
      } else {
        switch (thisClass,
            "numeric" = {thisPred[,thisVar] = median(df[,thisVar])},
            "factor" = {thisPred[,thisVar] = levels(df[,thisVar])[[1]]},
            "ordered" = {thisPred[,thisVar] = levels(df[,thisVar])[[1]]}
              )
      }
    }

    thisPred <- thisPred %>% select(-init)
    thisPred <- do.call("rbind", replicate(length(labs), thisPred, simplify = FALSE))

    pred <- cbind(labPred,thisPred)
    p<-data.frame(predict(modobj$gam,pred,se.fit = T))
    pred <- cbind(pred,p)
    pred$selo <- pred$fit - 2*pred$se.fit
    pred$sehi <- pred$fit + 2*pred$se.fit
    pred[,group_var] = NA
    pred[,thisResp] = 1

    low_color = "#91bfdb"
    high_color = "#fc8d59" 
    high_line = "#f46d43"
    low_line = "#4575b4"
    if (grad_fill == T) {
      p1 <- ggplot(data = df, aes_string(x = smooth_var,y = thisResp, color = int_var)) +
      geom_point(alpha = 0.85,stroke = 0, size = point_size) + geom_line(aes(group = bblid),alpha = .75 ) +
      scale_color_gradientn(colors = c(low_line,low_color,'grey90',high_color,high_line), limits = c(-2,2),values=c(0,.25,.5,.75,1),oob=squish,name = "") +
      geom_ribbon(data = pred,aes_string(x = smooth_var , ymin = "selo",ymax = "sehi", fill = "lab"),alpha = .18, linetype = 0) +
      scale_fill_manual(values = c(high_color,low_color)) +
      geom_line(data = pred,aes_string(x = smooth_var, y = "fit",group = "lab"),size = line_size) +
      labs(title = plabels, y = "R2* (1/sec)", x = "Age (years)", col = int_var)
    } else {

    p1 <- ggplot(data = df, aes_string(x = smooth_var,y = thisResp, color = int_var)) +
      geom_point(alpha = .35,stroke = 0, size = point_size) + geom_line(aes(group = bblid),alpha = .3) +
      scale_color_brewer(type = "qual",palette = 6,direction = -1) +
      geom_ribbon(data = pred,aes_string(x = smooth_var , ymin = "selo",ymax = "sehi", fill = int_var),alpha = .5, linetype = 0) +
      scale_fill_brewer(type = "qual",palette = 6,direction = -1) +
      geom_line(data = pred,aes_string(x = smooth_var, y = "fit",color = int_var),size = line_size) +
      scale_color_brewer(type = "qual",palette = 6,direction = -1) +
      labs(title = plabels, y = "R2* (1/sec)", x = "Age (years)")
    }
    if (str_detect(plabels,"Accumbens")) {
      p1 <- p1 + ylim(10,30)
    } else if (str_detect(plabels,"Pallidum")) {
      p1 <- p1 + ylim(15,30)
    } else if (any(str_detect(plabels,c("Putamen","Caudate")))) {
      p1 <- p1 + ylim(12,22)
    }
    p1<- p1+theme(text = element_text(size=font_size),
                axis.text = element_text(size = font_size,colour = "black"),
                legend.text = element_text(size = font_size),
                panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                panel.background = element_blank(),
                plot.background = element_blank())
    print(p1)
  } else {

  # line plot with no interaction
    thisPred <- data.frame(init = rep(0,np))

    for (v in c(1:length(theseVars))) {
      thisVar <- theseVars[[v]]
      thisClass <- varClasses[thisVar]
      if (thisVar == smooth_var) {
        thisPred[,smooth_var] = seq(min(df[,smooth_var],na.rm = T),max(df[,smooth_var],na.rm = T), length.out = np)
      } else {
        switch (thisClass,
            "numeric" = {thisPred[,thisVar] = median(df[,thisVar])},
            "factor" = {thisPred[,thisVar] = levels(df[,thisVar])[[1]]},
            "ordered" = {thisPred[,thisVar] = levels(df[,thisVar])[[1]]}
              )
      }
    }
    pred <- thisPred %>% select(-init)
    p<-data.frame(predict(modobj$gam,pred,se.fit = T))
    pred <- cbind(pred,p)
    pred$selo <- pred$fit - 2*pred$se.fit
    pred$sehi <- pred$fit + 2*pred$se.fit
    pred[,group_var] = NA
    pred[,thisResp] = 1

    pl <- ggplot(data = df, aes_string(x = smooth_var,y = thisResp)) +
      geom_point(alpha = .3,stroke = 0, size = point_size) + geom_line(aes(group = bblid),alpha = .3) +
      geom_ribbon(data = pred,aes_string(x = smooth_var , ymin = "selo",ymax = "sehi"),alpha = .5, linetype = 0) +
      geom_line(data = pred,aes_string(x = smooth_var, y = "fit"),size = line_size) +
      labs(title = plabels, y = "R2*", x = smooth_var)
    if (str_detect(plabels,"Accumbens")) {
      pl <- pl + ylim(10,30)
    } else if (str_detect(plabels,"Pallidum")) {
      pl <- pl + ylim(15,30)
    } else if (any(str_detect(plabels,c("Putamen","Caudate")))) {
      pl <- pl + ylim(12,22)
    }
    p1<-pl
  }
  
  if (check_diagnostics == T) {
    cp <- check(b,
    a.qq = list(method = "tnorm",
                a.cipoly = list(fill = "light blue")),
    a.respoi = list(size = 0.5),
    a.hist = list(bins = 10))
    print(cp)
  }

  return(p1)
}
```

```{r long vis, include = FALSE,fig.height=5,fig.width=3,warning=F}
# Special case for plotting age and age*sex models. Mostly redundant now that the above function exists, but it works and sets colors differently, so it is still used.

# *ALSO CALCULATES DERIVATIVES* 
longitudinal_plot <- function(modobj,plabels = NULL){
  df <- modobj$gam$model
  s<-summary(modobj$gam)
  thisResp <- as.character(modobj$gam$terms[[2]])

  pred.male <- data.frame(age = seq(min(df$age,na.rm = T),max(df$age,na.rm = T), length.out = 1000), oSex = 'male',sequence = "B0map_onesizefitsall_v3",timepoint = "1",visitnum = "1",oRace_bwo = "white")
  p<-data.frame(predict(modobj$gam,pred.male,se.fit = T))
  pred.male$fit = p$fit
  pred.male$selo = p$fit - 2*p$se.fit
  pred.male$sehi = p$fit + 2*p$se.fit
  pred.male[,thisResp] = 1
  pred.male$bblid=1
  
  pred.female <- data.frame(age = seq(min(df$age,na.rm = T),max(df$age,na.rm = T), length.out = 1000), oSex ='female',sequence = "B0map_onesizefitsall_v3",timepoint = "1",visitnum = "1",oRace_bwo = "white")
  p<-data.frame(predict(modobj$gam,pred.female,se.fit = T))
  pred.female$fit = p$fit
  pred.female$selo = p$fit - 2*p$se.fit
  pred.female$sehi = p$fit + 2*p$se.fit
  pred.female[,thisResp] = 1
  pred.female$bblid=1

  female <- "#e41a1c"
  male <- "#377eb8"
  # geom_point(alpha = .25) +
  p1 <- ggplot(data = df,aes_string(x = "age", y = thisResp, group = "bblid",color = "oSex",fill = "oSex")) +
    geom_line(alpha = .3) + geom_point(alpha = .3,stroke = 0, size = point_size,show.legend = F) +
    scale_color_manual(values = c(male,female)) +scale_fill_manual(values = c(male,female)) +
    geom_ribbon(data = pred.male, aes(ymin=selo, ymax=sehi), fill = male, alpha = .2, linetype = 0,show.legend = F) +
    geom_line(data = pred.male,aes(x = age, y = fit), color = male,size = line_size) +
    geom_ribbon(data = pred.female, aes(ymin=selo, ymax=sehi), fill = female, alpha = .2, linetype = 0,show.legend = F) +
    geom_line(data = pred.female,aes(x = age, y = fit), color = female,size = line_size) +
    labs(title = plabels, y = "R2* (1/sec)", x = "Age (years)", col = "Sex")  + ylim(min(pred.female$selo-.5),NA)+
    theme(text = element_text(size=font_size))
if (str_detect(plabels,"Accumbens")) {
    p1 <- p1 + ylim(10,30)
  } else if (str_detect(plabels,"Pallidum")) {
    p1 <- p1 + ylim(10,30)
  } else if (any(str_detect(plabels,c("Putamen","Caudate")))) {
    p1 <- p1 + ylim(12,22)
  }

p1<- p1+theme(text = element_text(size=font_size),
              axis.text = element_text(size = font_size),
              axis.title.x = element_blank(),
              axis.text.x = element_blank(),
              axis.ticks.x = element_blank(),
              legend.text = element_text(size = font_size),
              axis.title = element_text(size = font_size),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              panel.background = element_rect(fill = "transparent",colour = NA),
              plot.background = element_rect(fill = "transparent",colour = NA),
              plot.margin = unit(c(.2, .2, 0, .2), "cm")) #Top, left,Bottom, right

#### Add derivative plot
# function to extract derivative, confidence interval, significance, and plot
get_derivs_and_plot <- function(modobj,low_color=NULL,hi_color=NULL){
  if (is.null(low_color)){low_color = "white"}
  if (is.null(hi_color)){hi_color = "grey20"}
  derf<-derivatives(modobj)
  derf<- derf %>%
    mutate(sig = !(0 >lower & 0 < upper))
  derf$sig_deriv = derf$derivative*derf$sig
  cat(sprintf("\nSig change: %1.2f - %1.2f\n",min(derf$data[derf$sig==T]),max(derf$data[derf$sig==T])))
  d1<- ggplot(data=derf) + geom_tile(aes(x = data, y = .5, fill = sig_deriv)) + 
    scale_fill_gradient(low = low_color,high = hi_color,limits = c(0,.5),oob = squish)+ #Holding the limit constant across all plots (at .5)
    labs(x = "Age (years)",fill = "\u0394R2*/Year") + 
    theme(axis.title.y = element_blank(),
          axis.text.y = element_blank(),
          axis.text.x = element_text(size = font_size),
          axis.line = element_blank(),
          axis.ticks.y = element_blank(),
          text = element_text(size=font_size),
          legend.text = element_text(size = font_size),
          axis.title = element_text(size = font_size),
          legend.key.width = unit(.5,"cm"),
          legend.position = "right",
          plot.margin = unit(c(0, 0, .2, 0), "cm"))+
    guides(fill = guide_colorbar(reverse = F,direction = "horizontal",title.position = "top")) +
    geom_rect(aes(ymin=0,ymax=1,xmin=min(data),xmax=max(data)),color="black",fill="white",alpha = 0)
  return(d1)
}

## NOTE: Right now, derivative calculation is not working for by= factor models in gratia
##       For interaction models, to get significant derivatives for male and female separately, we need to fit the model for male and females separately and save the derivatives.

if (any(grep(x = row.names(s$s.table),pattern =  ":"))) {
  # Male and female separately if there is an interaction in the model.
  f<-formula(modobj$gam) # current formula
  fterms <- terms(f)
  fac <- attr(fterms, "factors")
  idx <- which(as.logical(colSums(fac[grep(x=row.names(fac),pattern = "oSex"),])))
  new_fterms <- drop.terms(fterms, dropx = idx, keep.response = TRUE)
  new_formula <- formula(new_fterms) # Formula without any sex terms in the model.
  malemod <- gamm(formula = new_formula,data = df,subset = oSex == "male",random=list(bblid=~1))
  femalemod <- gamm(formula = new_formula,data = df,subset = oSex == "female",random=list(bblid=~1))
  d1_male <- get_derivs_and_plot(modobj = malemod,low_color = "white",hi_color = male)
  d1_female <- get_derivs_and_plot(modobj = femalemod,low_color = "white",hi_color = female)
  #This one will be on top, so we don't need any labels, etc.
  d1_male$theme$axis.title = element_blank()
  d1_male$theme$axis.text.x = element_blank()
  d1_male$theme$axis.ticks=element_blank()
  d1_male$theme$plot.margin = unit(c(0,0,0,0),"cm")
  d1_male$labels$fill=NULL
  d1_female$labels$fill=NULL
  scatter <- list(p1)
  bar1 <- list(d1_male)
  bar2 <- list(d1_female)
  allplots <- c(scatter,bar1,bar2)
  pg<-plot_grid(rel_heights = c(24,1,3),plotlist = allplots,align = "v",axis = "lr",ncol = 1)
} else {
  # No need to split
  d1 <- get_derivs_and_plot(modobj = modobj)
  scatter <- list(p1)
  bar <- list(d1)
  allplots <- c(scatter,bar)
  pg<-plot_grid(rel_heights = c(8,1),plotlist = allplots,align = "v",axis = "lr",ncol = 1)
}

print(pg)
ggsave(plot = pg,filename = gsub(x = sprintf("line_deriv_plot_%s.svg",plabels),pattern = " ",replacement = "_"),device = "svg",width = 220,height = 180,units = "mm")
return(pg)
}
```


# Load the data
```{r load data, warning=FALSE, include=T }
# Load the datafile
dataFile <- "brain_iron_data.Rda"
dataTable <- readRDS(file=dataFile)

## Create the final sample
### Remove
### 1. Projects with incompatible data
dataTable <- dataTable %>%
  filter(bblid != 110689)%>%
  filter(!(ProjectName %in% c("DAY2_808799",
                              "FNDM1_810211",
                              "FNDM2_810211",
                              "NEFF_V2",
                              "NEFF_PILOT"
                              )
           )
         ) #Not using data from these projects or participants with no scans.
### 2. Sequences with odd parameters
dataTable <- dataTable %>%
  filter(sequence != 'B0map_v4_matchedFOV' & sequence != "B0map") %>% # Different voxel sizes
  filter(sequence != 'B0map_onesizefitsall_v3_T2S') %>% # small number of scans with odd params
  filter(!is.na(GOOD)&!is.na(Putamen)) # No r2* data available or no QC possible
#Print some info
summaryTable<-dataTable%>%group_by(bblid,sex)%>%summarize(n=n())
cat(sprintf('Initial sample includes:\n %d individuals aged %1.2f - %1.2f (M = %1.3f, SD = %.3f)\n (M/F = %d/%d)\n%d observations', 
        length(unique(dataTable$bblid)),
        min(dataTable$age,na.rm=T),
        max(dataTable$age[dataTable$visit==1],na.rm=T),
        mean(dataTable$age[dataTable$visit==1],na.rm=T),
        sd(dataTable$age[dataTable$visit==1],na.rm=T),
        sum(summaryTable$sex=='male'),
        sum(summaryTable$sex=='female'),
        length(dataTable$bblid)))
### 3. Subjects that don't meet inclusion
cat(sprintf('\nHealth exclusions: %d subs, %d scans',
            length(unique(dataTable$bblid[dataTable$ltnExcludev2==1])),
            sum(dataTable$ltnExcludev2==1,na.rm = T)))
dataTable <- dataTable %>% 
    filter(ltnExcludev2 ==0 | is.na(ltnExcludev2)) # exclude unhealthy/medicated subs
### 4. Scans that don't pass QA
cat(sprintf('\nQC exclusions: %d scans fail (%d pass)',
            sum(dataTable$GOOD==0,na.rm = T),
            sum(dataTable$GOOD %in% c("1","2"),na.rm = T)))
dataTable <- dataTable %>%
  filter(GOOD %in% c("1","2")) # keep only subjects that pass QA
dataTable$acquisition = factor(dataTable$sequence,levels = c("B0map_onesizefitsall_v3", "B0map_v4","B0map_onesizefitsall_v2","b0map_v4"), labels = c("A","B","C","D"))

## Count visits  
dataTable <- dataTable %>% 
  group_by(bblid) %>% 
  mutate(visitnum = min_rank(ScanAgeMonths)) %>%
  ungroup()
  dataTable$visitnum <- ordered(dataTable$visitnum)
summaryTable<-dataTable%>%group_by(bblid,sex)%>%summarize(n=n())
cat(sprintf('\nFinal sample includes:\n %d individuals aged %1.2f - %1.2f (M = %1.3f, SD = %.3f)\n (M/F = %d/%d)\n %d observations', 
        length(unique(summaryTable$bblid)),
        min(dataTable$age,na.rm=T),
        max(dataTable$age[dataTable$visit==1],na.rm=T),
        mean(dataTable$age[dataTable$visit==1],na.rm=T),
        sd(dataTable$age[dataTable$visit==1],na.rm=T),
        sum(summaryTable$sex=='male'),
        sum(summaryTable$sex=='female'),
        length(dataTable$bblid)))

```

```{r figure 1}
# Create sample description figure (Fig 1)
sortedBBLID<-dataTable %>% group_by(bblid) %>% summarise(m=min(age)) %>% arrange(m)
sortedBBLID$row<-as.numeric(rownames(sortedBBLID))
newDataTable<-dataTable%>%left_join(sortedBBLID%>%select(bblid,row),by="bblid")
fig1<-ggplot(data = newDataTable,aes(x=reorder(row,age,FUN = min),y=age,color=oSex)) + 
  geom_point(size=.5,alpha=.5) + geom_line(alpha=.5) +
  scale_x_discrete(breaks = c(1,500,1000,length(sortedBBLID$bblid))) +
  coord_flip(clip = "off") + 
  scale_color_manual(values = c("blue","red"),labels=c("Male","Female"))+ 
  labs(y = "Age (years)",x="Participant",color="") +theme(legend.position = c(.8,.2))
print(fig1)
ggsave("Fig1.svg",device = "svg",plot = fig1,width = 85,height = 85,units = "mm",bg = "transparent")

fig1b <- ggplot(data = dataTable, aes(x = age, color = acquisition, fill = acquisition)) + geom_histogram(alpha = .8) + scale_fill_brewer(palette = "Set1") + scale_color_brewer(palette = "Set1") + xlab("Age") + ylab("Number of Participants") + labs(fill = "Acquisition", color = "Acquisition") + scale_x_continuous(breaks = c(8,12,16,20,24)) + theme(legend.position = c(.8,.8))
ggsave(plot=fig1b,"Fig1b.svg",device = "svg",width = 85,height = 100,units = "mm",bg = "transparent")

```

```{r cognition_sample}
## make data frame for participants that have cognition data.
behTable <- dataTable %>% 
  filter(!is.na(NAR_Overall_Efficiency)) %>%
  filter(scan2cnbmonths<=6)
summaryTable<-behTable%>%group_by(bblid,sex)%>%summarize(n=n())
cat(sprintf('\nBehavior sample includes:\n %d individuals aged %1.2f - %1.2f (M = %1.3f, SD = %.3f)\n (M/F = %d/%d)\n %d observations', 
        length(unique(summaryTable$bblid)),
        min(behTable$age,na.rm=T),
        max(behTable$age[behTable$visit==1],na.rm=T),
        mean(behTable$age[behTable$visit==1],na.rm=T),
        sd(behTable$age[behTable$visit==1],na.rm=T),
        sum(summaryTable$sex=='male'),
        sum(summaryTable$sex=='female'),
        length(behTable$bblid)))
```

### Combat harmonization
```{r combat}
## Combat harmonization
# First harmonize for age models
comtable <- dataTable %>%
  select(Caudate,Putamen,Accumbens_Area,Pallidum,sequence,age,oSex,bblid,scanid,visitnum) %>%
  drop_na()
batch <- as.character(comtable$sequence)
ctab <- t(data.matrix(comtable%>%select(Caudate,Putamen,Accumbens_Area,Pallidum)))
g<-gamm(Putamen ~ visitnum + s(age, k = 4, fx = T) + oSex + s(age, by = oSex, k = 4, fx = T) ,data=comtable,random=list(bblid=~1))
mod<- model.matrix(g$gam)
combatdata <- combat(ctab,batch,mod=mod, eb = F)
harmonized_data<-data.frame(t(combatdata$dat.combat))
colnames(harmonized_data)<- paste(colnames(harmonized_data),"h",sep = "_")
harmonized_data$bblid=comtable$bblid
harmonized_data$scanid=comtable$scanid
dataTable<- dataTable %>%
  left_join(harmonized_data,by = c("bblid","scanid"))

# Now for Behavior models
comtable <- behTable %>%
  select(Caudate,Putamen,Accumbens_Area,Pallidum,sequence,age,NAR_Overall_Efficiency,oSex,bblid,scanid) %>%
  drop_na()
batch <- as.character(comtable$sequence)
ctab <- t(data.matrix(comtable%>%select(Caudate,Putamen,Accumbens_Area,Pallidum)))
g<-gamm(Putamen ~ s(age, k = 4, fx = T) + s(age, by = NAR_Overall_Efficiency, k = 4, fx = T) + oSex,data=comtable,random=list(bblid=~1))
mod<- model.matrix(g$gam)
combatdata <- combat(ctab,batch,mod=mod, eb = F)
harmonized_data<-data.frame(t(combatdata$dat.combat))
colnames(harmonized_data)<- paste(colnames(harmonized_data),"h",sep = "_")
harmonized_data$bblid=comtable$bblid
harmonized_data$scanid=comtable$scanid
behTable<- behTable %>%
  left_join(harmonized_data,by = c("bblid","scanid"))

longTable <- dataTable %>%
    gather(key = "ROI",value="R2s",Accumbens_Area_h,Putamen_h,Caudate_h,Pallidum_h)

rois <- c("Caudate_h","Putamen_h","Accumbens_Area_h","Pallidum_h")
```

---
# ROI comparison
Analyses for figure 3
```{r roicompare,eval=T,results="asis"}
source("R_rainclouds.R") #requires RainCloud package: https://github.com/RainCloudPlots/RainCloudPlots/blob/master/tutorial_R/R_rainclouds.R

longTable$ROI<-factor(longTable$ROI,levels = c("Caudate_h","Putamen_h","Accumbens_Area_h","Pallidum_h"),labels = c("Caudate","Putamen","Nucleus Accumbens","Pallidum"))
longTable$oldROI <- c("Caudate","Putamen","Accumbens_Area","Pallidum")

longTable$num_voxels[longTable$ROI=="Caudate"] <- 132
longTable$num_voxels[longTable$ROI=="Putamen"] <- 175
longTable$num_voxels[longTable$ROI=="Nucleus Accumbens"] <- 19
longTable$num_voxels[longTable$ROI=="Pallidum"] <- 58

l<-lmerTest::lmer(R2s ~ sex + age + ROI + (1|bblid),data = longTable)
m<-psycho::get_means(fit = l,formula="ROI")
emmeans_table<- kable(m) %>% 
  kable_styling(bootstrap_options = "striped",position = "left")
print(emmeans_table)

c <- psycho::get_contrasts(l,"ROI",adjust="bonf",effsize = "cohen1998")
contrast_table<- kable(c) %>% 
  kable_styling(bootstrap_options = "striped",position = "left")
print(contrast_table)
write.csv(c,'EMMeans_contrast.csv',row.names = F)

fig3<-ggplot(data=longTable,aes(x = reorder(ROI,-R2s,FUN = mean,na.rm=T),y = R2s,fill=ROI,color = ROI)) + 
  geom_flat_violin(show.legend = F,position=position_nudge(x=.2),alpha = .4) + 
  labs(x=NULL,y = "R2* (1/sec)",font_size=font_size) +
  geom_point(size=.25,alpha=.4,position = position_jitter(width = .1),show.legend = F) +
  geom_boxplot(color = "black",fill=NA,outlier.shape = NA,width=.1,show.legend = F) +
  scale_color_brewer(palette = "Set1") + scale_fill_brewer(palette = "Set1") +
  scale_x_discrete(labels=c("Globus\nPallidus","Nucleus\nAccumbens","Putamen","Caudate"))
fig3
ggsave(plot=fig3,"Fig3.svg",device = "svg",width = 85,height = 85,units = "mm",bg = "transparent")
```

# Developmental effects
Analyses for figure 4.  
Fit nonlinear age effect and check for age*sex interaction (factor smooth interaction: `s(age, by = oSex)`)
oSex is Sex specified as an ordered factor to appropriately test for the interaction (female against male reference)
```{r developmental, results="asis",eval=T,warning=F,fig.height=8,fig.width=8}
model_formula <- "R2s ~ visitnum +oSex +s(age, k = 4, fx = T) + s(age, by = oSex, k = 4, fx = T) "
## This will perform model testing. If the age * sex interaction is significant, it is retained, otherwise it will return an age + sex model.
models <- longTable %>%
    group_by(ROI)%>%
    nest()%>%
    mutate(results=purrr::pmap(list(data,model_formula,this_label = ROI, pbootstrap = T, longPlot = T),.f=gamm_model))
```

## Modeling of continuous*continuous interactions tests two possible models.  
Analyses for figures 5 and 6.

We compare an additive (main effects + interaction) bivariate smooth model and a varying coefficient model (details below).
The best model is chosen via the smallest BIC.
After the best interaction  model is selected, the significance of the interaction is tested with a parametric bootstrap likelihood ratio test. This test compares the model with the interaction term against a simpler nested model with main effects only. If the interaction model is significantly better, we keep that model. If not, the final model is the simpler model with no interaction.  

### 2D smooth with additional main effects and tensor product smooth, `ti`: `ti(age) + ti(Cognition) + ti(age,Cognition)`

From documentation: 
This model specifies a main effects + interaction structure such as:
`y ~ ti(x) + ti(z) + ti(x,z)`  

`ti` is the proper way of specifying an interaction term in the context of included main effect terms:  

"This functional ANOVA decomposition is supported by ti terms, which produce tensor product interactions from which the main effects have been excluded, under the assumption that they will be included separately. For example the ~ ti(x) + ti(z) + ti(x,z) would produce the above main effects + interaction structure. This is much better than attempting the same thing with s or te terms representing the interactions (although mgcv does not forbid it). Technically ti terms are very simple: they simply construct tensor product bases from marginal smooths to which identifiability constraints (usually sum-to-zero) have already been applied: correct nesting is then automatic (as with all interactions in a GLM framework). See Wood (2017, section 5.6.3)."  

### Varying coefficient model (using `by =`)

This will make the fit linear (rather than non-linear smooth) in the `by` variable (which is cognition in this case).
From documentation: 
"When using `by` with a numberic covariate, "the by argument ensures that the smooth function gets multiplied by covariate z"

```{r Cognitive, results="asis",warning=F, eval = T}

# Loop over ROIs
for (r in rois) {
  cat(sprintf("\n## Results for %s\n",r))
  # Set the cognitive variables to test
  if (r == "Putamen_h") {
    # Overall efficiency is significant for Putamen, so check the sub-domains of efficiency as well.
    cog_vars <- c("NAR_Overall_Efficiency",
                  "NAR_F1_Social_Cognition_Efficiency",
                  "NAR_F2_Complex_Reasoning_Efficiency",
                  "NAR_F3_Executive_Efficiency",
                  "NAR_F4_Memory_Efficiency")
  } else{
    cog_vars <- c("NAR_Overall_Efficiency")
  }
  # Loop over the cognitive variables
  for (cv in cog_vars) {
    cat(sprintf("\n### %s\n",cv))
    thisTable <- behTable
    
    # First look at a main effects only model
    # Fitting smooths for age and cognition
    add_formula <- sprintf("%s ~ timepoint +oSex + s(age, k=4, fx = T) + s(%s, k=4, fx = T)",r,cv,cv)
    gamm_model(thisTable,
               add_formula,
               this_label = sprintf("%s %s M.E.",r,cv),
               smooth_var = cv,
               group_var = "bblid",
               pbootstrap = T,
               model_test = F)
    
    # Compare the two interaction models
    cat('\n### Comparing interaction models...\n')
    # Bivariate interaction
    bv_formula <- sprintf(
      "%s ~ timepoint +oSex + ti(age, k=4, fx = T) + ti(%s, k=4, fx = T) + ti(age,%s, k=4, fx = T)",
      r,cv,cv,cv)
    # Linear varying coefficient interaction
    vc_formula <- sprintf(
      "%s ~ timepoint + oSex +s(age, k=4, fx = T) + s(age, by = %s, k=4, fx = T)",
      r,cv)

    bv <- gamm(as.formula(bv_formula),
               random = list(bblid=~1),
               data = thisTable)
    vc <- gamm(as.formula(vc_formula),
               random = list(bblid=~1),
               data = thisTable)
    bic<-BIC(bv$lme,vc$lme) # get BIC
    bestmod <- gsub(row.names(bic)[which.min(bic$BIC)],pattern = "$lme",replacement = "", fixed = T) #best is min BIC
    
    #confirm there are no concurvity issues if the VC model is the best.
    c<-as.data.frame(concurvity(vc$gam))
    if (c["observed",contains(vars = names(c),"s(age):")]>.75) {
      bestmod<-"bv"
    }
    switch (bestmod,
      "bv" = {model <- bv},
      "vc" = {model <- vc}
    )

    model_formula <- model$gam$formula
    cat("\n\nbest model is\n",deparse(model_formula))

    # Now check if the interaction is significant using parametric bootstrap `(pbootstrap=T)`
    gamm_model(thisTable,
               model_formula,
               this_label = sprintf("%s %s final model",r,cv),
               smooth_var = "age",
               int_var = cv,
               group_var = "bblid",
               pbootstrap = T,
               model_test = F)

  }
}
```

```{r 3-way interaction, results="asis",warning=F, eval = F}
cog_vars <- c("NAR_Overall_Efficiency")

for (r in rois) {
  cat(sprintf("\n## Results for %s\n",r))
  for (cv in cog_vars) {
    cat(sprintf("\n### %s\n",cv))
    thisTable <- dataTable

    ## The logic here is to compare a bivariate interaction that varies by sex to a nested model that does not vary by sex
    cat('\n### Comparing interaction models...\n')
    # Bivariate interaction that just includes sex as a covariate
    bv1_formula <- sprintf(
      "%s ~ timepoint + oSex + te(age,%s, k=4, fx = T)",
      r,cv,cv,cv)
    # Bivariate that varies as a function of sex
    bv2_formula <- sprintf(
      "%s ~ timepoint + oSex + te(age,%s, k=4, fx = T) + te(age,%s, by = oSex, k=4, fx = T)",
      r,cv,cv)
  
    bv1 <- gamm(as.formula(bv1_formula),
               random = list(bblid=~1),
               data = thisTable)
    bv2 <- gamm(as.formula(bv2_formula),
               random = list(bblid=~1),
               data = thisTable)
    bic<-BIC(bv1$lme,bv2$lme) # get BIC
    bestmod <- gsub(row.names(bic)[which.min(bic$BIC)],pattern = "$lme",replacement = "", fixed = T) #best is min BIC
    bestmod<-"bv2"
    switch (bestmod,
      "bv1" = {model <- bv1},
      "bv2" = {model <- bv2}
    )
    model_formula <- model$gam$formula
    cat(sprintf("\n\nbest model is %s\n",deparse(model_formula)))
    a <- anova(model$gam)
    print(kable(a$pTerms.table)%>%kable_styling(position = "left"))
    print(kable(a$s.table)%>%kable_styling(position = "left"))
    
    # Now check if the interaction term from the chosen model is significant
    gamm_model(thisTable,
               model_formula,
               this_label = sprintf("%s %s final model",r,cv),
               smooth_var = "age",
               int_var = cv,
               group_var = "bblid",
               pbootstrap = F,
               model_test = F)
  }
}

```


## Controlling for sequence

Though data are harmonized w.r.t sequence, reviewer asked to confirm sequence does not impact results.

We will test this for development and for age*Overall cogntion models.

```{r sequence, results="asis",warning=F, eval = T}
# Developmental model
dev_formula <- "Putamen_h ~ visitnum +oSex + acquisition +s(age, k = 4, fx = T) "
g <- gamm(as.formula(dev_formula),
           random = list(bblid=~1),
           data = behTable)
anova(g$gam)

# Cognitive interaction
cog_formula <- "Putamen_h ~ timepoint + oSex + sequence +s(age, k=4, fx = T) + s(age, by = NAR_Overall_Efficiency, k=4, fx = T)"
cog <- gamm(as.formula(cog_formula),
           random = list(bblid=~1),
           data = behTable)
anova(cog$gam)

```